---
title: "Half-lives"
format: html
editor: visual
---

```{r}
#| echo: false
#| include: false
#| warning: false

library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
library(kableExtra)
library(ggpubr)
library(plotly)
```

```{r}
#| echo: false
#| include: false

# This is just a filtered by Post-translational
human_ptms_hl <- read.csv(file = "/Users/anastasialinchik/Desktop/Thesis/DATA/generated/human_ptms_hl.tsv", header = T, sep = '\t')
cell_types_hl <- read.csv(file = '/Users/anastasialinchik/Desktop/Thesis/DATA/generated/long_proteins_hl_cell_types.tsv', header = T, sep = '\t')
long_proteins_hl <- read.csv(file = '/Users/anastasialinchik/Desktop/Thesis/DATA/generated/long_proteins_hl_cell_types.tsv', header = T, sep = '\t')
ageing_proteins <- read_excel( '/Users/anastasialinchik/Desktop/Thesis/DATA/Age-correlated protein.xlsx') %>% filter(Animal == 'Human') %>% dplyr::select(`Uniprot entry`)
ageing_proteins <- human_ptms_hl %>% filter(LeadProt %in% ageing_proteins$`Uniprot entry`)
oxPTMs <- read.table(file = '/Users/anastasialinchik/Desktop/Thesis/Master_thesis/oxPTMs_unimod.txt', header = T) 
hl <- read.csv(file = '/Users/anastasialinchik/Desktop/Thesis/DATA/generated/hl.tsv', header = T, sep = '\t')
genage <- read.csv(file = '/Users/anastasialinchik/Desktop/Thesis/DATA/genage_human.csv', header = T)
```

## PTMs and protein half-lives

-   The half-life of a protein is the time it takes for the concentration of a radioactively protein to decrease by a half. Protein half-lives can be used as estimates of residence time of proteins in the cell.

-   Proteins that reside longer in the cell may be more susceptible to oxidative damage.

-   Protein half-lives were obtained from: [paper1](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5975921/), [paper2](https://www-science-org.kuleuven.e-bronnen.be/doi/10.1126/science.1199784#appendix).

-   It is assumed that each protein only has one modification. The proteins with no modifications are not identified.

### Distribution of protein half-lives

Approach:

-   Get a list of PTMs that correlate with ageing such as oxPTMs, acetylation etc.

-   Test whether the abundance of these PTMs changes between the long-lived proteins and the normal proteins.

Below is a comparison of the distribution of the half-lives that was found in literature and the distribution of a subset of those half-lives in the proteins found in the dataset.

```{r}
#| echo: false

ggplot() +
  geom_density(data = hl, aes(x = mean_hl_hours, fill = 'Half-lives in papers'), alpha = 0.7, bw = 0.4) +
  geom_density(data = human_ptms_expanded_hl, aes(x = mean_hl_hours, fill = 'Proteins in the dataset'), alpha = 0.7, bw = 0.4) + 
  labs(title = 'Distributions of protein half-lives', x = 'Mean half-lives (hours)', y = 'Density') +
  theme_classic() +
  scale_fill_manual(values = c("#5DB7B1", '#FB8500'))  # Manually specify fill colors
```

-   There is an enrichment of proteins with short half-lives and proteins with half-lives just below 10 hours.

The weighted mean half-life of proteins in the dataset is 8.37 hours, which is within the expected range when compared to values found in [literature](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3788602/).

```{r}
#| echo: false
#| include: false

w_mean <- weighted.mean(human_ptms_hl$mean_hl_hours, human_ptms_hl$norm_counts)
w_mean
```

A boxplot was plotted in order to better visualise the proteins with exceptionally high half-lives. The boxplot does not use normalised counts due to the high number of data points. It can be seen that there are outliers in the boxplot, most of which have a high mean half-life. In total, there are 84083 outliers.

```{r}
#| echo: false
ggplot(human_ptms_hl, aes(y = mean_hl_hours)) +
  geom_boxplot() +
  theme_classic()
```

-   The distribution of the half-lives of the outliers is plotted below. Here, normalised counts are used.

```{r}
#| echo: false

# here the weights are not taken into account when calculating outliers

# Calculate boxplot statistics
boxplot_stats <- boxplot.stats(human_ptms_expanded_hl$mean_hl_hours)

# Calculate IQR
IQR <- boxplot_stats$stats[4] - boxplot_stats$stats[2]

# Define lower and upper bounds for outliers
lower_bound <- boxplot_stats$stats[2] - 1.5 * IQR
upper_bound <- boxplot_stats$stats[4] + 1.5 * IQR

# Filter outliers
outliers_df <- human_ptms_expanded_hl %>%
  filter(mean_hl_hours < lower_bound | mean_hl_hours > upper_bound)

ggplot(outliers_df, aes(y = mean_hl_hours, weight = norm_counts)) +
  geom_boxplot() + 
  labs(title = 'Distribution of half-lives among outlier proteins', y = 'Mean half-lives (hours)') +
  theme_classic()
```

Though the mean half-life of these proteins is higher than the mean half-life of the whole dataset, none of the outlier proteins can be classified as long-lived proteins. Proteins can be classified as long-lived when their mean half-life exceeds 48 hours ([ref](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5500981/#:~:text=Many%20of%20your%20body's%20proteins,%2Dlived%20protein%20(LLP).)), though this is an arbitrary definition. Nonetheless, the change in PTM abundances can still be investigated.

```{r}
#| echo: false
#| include: false

# Violin plot for oxPTMs
human_ptms_oxPTMs <- human_ptms_hl %>% filter(unimod_id %in% oxPTMs$ID)

ggplot(human_ptms_oxPTMs, aes(x = '', y = mean_hl_hours, weight = norm_counts)) +
  geom_violin(bw=0.5, fill = '#73BFB8') +
  labs(x = 'oxPTMs', y = 'Half-life (hours)', title = 'Distribution of protein half-lives separated by PTMs') +
  geom_abline(intercept = 2.3, slope = 0, colour = 'red') +
  theme_classic(base_size = 12)
```

-   A similar distribution is seen as before. Two peaks can be clearly seen here. It would be interesting to look at the difference between the proteins present in the two peaks. The red line is used to split the proteins into two groups.
-   Is the small peak due to the distribution of the half-lives found in various papers? This can be checked using the plot below.
-   The plot below has only one peak, meaning that small peak previously seen in the violin plot is not due to the distribution of the original half-lives.

Below are the proteins that are present in the first peak in the violin plot.

```{r}
#| echo: false
#| eval: true

short_hl_proteins <- human_ptms_hl %>% filter(mean_hl_hours < 2.3) %>% dplyr::select(LeadProt) %>%
  distinct()

kable(short_hl_proteins)
```

### Phosphorylation

-   Only the modification \[21\]Phospho is present here.

Splitting the dataset in a group with phosphorylation proteins and another group with all remaining proteins.

```{r}
human_ptms_hl_mod_group <- human_ptms_hl
human_ptms_hl_mod_group$mod_group <- if_else(human_ptms_hl$unimod_id == 21, 'Phospho', 'Non-Phospho')


ggplot(human_ptms_hl_mod_group, aes(x = mean_hl_hours, weight = norm_counts, fill = mod_group)) +
  geom_density(alpha = 0.6, bw = 0.4) + 
  labs(title = 'Distribution of half lives of phosphorylated proteins', x = 'Mean half-life (hours)', y = 'Counts')+
  theme_classic()
```

Distribution of phosphorylated ageing proteins vs phosphorylated non-ageing proteins vs non-phosphorylated proteins

```{r}
pho_age_prot <- ageing_proteins %>% 
  filter(unimod_id == 21) 

pho_age_prot <- human_ptms_hl %>% 
  filter(unimod_id == 21) %>%
  filter(Uniprot_entry_name %in% genage$uniprot)

pho_no_age_prot <- human_ptms_hl %>% filter(unimod_id == 21) %>%
  filter(!LeadProt %in% pho_age_prot$LeadProt)

no_pho_prot <-  human_ptms_hl %>% filter(!unimod_id == 21) %>%
  filter(!LeadProt %in% pho_age_prot$LeadProt)

ggplot() +
  geom_density(data = no_pho_prot, aes(x = mean_hl_hours, fill = 'no_pho_prot'), alpha = 0.7, bw = 0.4) +
  geom_density(data = pho_no_age_prot, aes(x = mean_hl_hours, fill = 'pho_no_age_prot'), alpha = 0.7, bw = 0.4) +
   geom_density(data = pho_age_prot, aes(x = mean_hl_hours, fill = 'pho_age_prot'), alpha = 0.7, bw = 0.6) +
  labs(title = 'Distributions of protein half-lives', x = 'Mean half-lives (hours)', y = 'Density') +
  theme_classic() +
  scale_fill_manual(values = c('no_pho_prot' = '#FB8500', 'pho_no_age_prot' = '#2364AA','pho_age_prot'  = "#5DB7B1"), name = 'Legend', breaks = c('no_pho_prot', 'pho_no_age_prot','pho_age_prot' ))  # Manually specify fill colors
```

### Methylation

-   Filtered by the \[34\]Methyl modification

```{r}
human_ptms_hl_mod_group$mod_group <- if_else(human_ptms_hl$unimod_id == 34, 'Methyl', 'Non-Methyl')

ggplot(human_ptms_hl_mod_group, aes(x = mean_hl_hours, weight = norm_counts, fill = mod_group)) +
  geom_density(alpha = 0.6, bw = 0.4) + 
  labs(title = 'Distribution of half lives of phosphorylated proteins', x = 'Mean half-life (hours)', y = 'Counts')+
  theme_classic()
```

### Acetylation

-   Filtered by the \[1\]Acetyl modification

```{r}
#| echo: false
human_ptms_hl_mod_group$mod_group <- if_else(human_ptms_hl$unimod_id == 1, 'Acetyl', 'Non-Acetyl')

ggplot(human_ptms_hl_mod_group, aes(x = mean_hl_hours, weight = norm_counts, fill = mod_group)) +
  geom_density(alpha = 0.6, bw = 0.4) + 
  labs(title = 'Distribution of half lives of phosphorylated proteins', x = 'Mean half-life (hours)', y = 'Counts')+
  theme_classic()
```

### Oxidation

-   Only modification: \[35\]Oxidation

```{r}
#| echo: false
human_ptms_expanded_hl_oxi <- human_ptms_hl %>% filter(ptm_name == '[35]Oxidation')
ggplot(human_ptms_expanded_hl_oxi, aes(x = mean_hl_hours, weight = norm_counts)) +
  geom_histogram(bins = 50) +
  labs(title = 'Distribution of half-lives of oxidised proteins', x = 'Mean half-life (hours)', y = 'Counts')+
  theme_classic()
```

```{r}
#| echo: false

human_ptms_hl_mod_group$mod_group <- if_else(human_ptms_hl$unimod_id == 35, 'Oxidation', 'Non-Oxidation')

ggplot(human_ptms_hl_mod_group, aes(x = mean_hl_hours, weight = norm_counts, fill = mod_group)) +
  geom_density(alpha = 0.6, bw = 0.4) + 
  labs(title = 'Distribution of half lives of oxidised proteins', x = 'Mean half-life (hours)', y = 'Counts')+
  theme_classic()
```

### oxPTMs

All PTMs related to oxidative damage in general, not only \[35\]Oxidation.

```{r}
#| echo: false
human_ptms_hl_oxi <- human_ptms_hl %>% filter(unimod_id %in% oxPTMs$ID)
ggplot(human_ptms_hl_oxi, aes(x = mean_hl_hours, weight = norm_counts)) +
  geom_histogram(bins = 50) +
  labs(title = 'Distribution of half-lives of oxidised proteins', x = 'Mean half-life (hours)', y = 'Counts')+
  theme_classic()
```

```{r}
#| echo: false
human_ptms_hl_mod_group$mod_group <- if_else(human_ptms_hl$unimod_id %in% oxPTMs$ID, 'oxPTMs', 'Non-oxPTMs')

ggplot(human_ptms_hl_mod_group, aes(x = mean_hl_hours, weight = norm_counts, fill = mod_group)) +
  geom_density(alpha = 0.5, bw = 0.4) +
  labs(title = 'Distribution of half-lives of oxidised proteins', x = 'Mean half-life (hours)', y = 'Counts')+
  theme_classic()
```

```{r}
#| echo: false

df <- human_ptms_hl_mod_group %>% dplyr::select(LeadProt, modification, mod_group, unimod_id, mean_hl_hours) %>%
  distinct()

ggplot(df, aes(x = mod_group, y = mean_hl_hours)) +
  geom_violin(bw=0.4, fill = '#73BFB8') +
  geom_boxplot(width=0.3, outlier.shape = NA) + 
  labs(x = 'PTM name', y = 'Half-life (hours)', title = 'Distribution of protein half-lives separated by PTMs') +
  stat_summary(fun=mean, geom="point", shape=19, size=2.5, colour = '#EA7317') +
  theme_classic(base_size = 12)
```

As can be seen in the above graphs, the presence of the small peak, which represents proteins with very short half-lives, varies depending on the type of modification. It is more prominent in PTMs that are related to ageing.

**Hypothesis**: proteins with higher mean half-lives remain in the cell for longer, therefore they are more susceptible to oxidative damage and will accumulate more oxPTMs.

Approach:

-   Count the total number of oxPTMs on each protein (using normalised counts)

-   Plot the mean half-life of the proteins vs their total oxPTMs count

Below is a plot that shows how the number of oxPTMs changes with the mean half-life of different proteins. The proteins, which were identified to have a very short half-life are shown in blue. A similar pattern is seen between the blue and red points, where some have a very high number of modifications, while most show very low abundances. Most proteins still seem to remain largely unmodified irrespective of their half-life.

```{r}
#| echo: false
#| warning: false
oxPTMs_hl <- human_ptms_hl %>% 
  inner_join(oxPTMs, by = join_by(unimod_id == ID)) %>%
  group_by(LeadProt) %>%
  summarise(mean_hl_hours, sum_oxPTMs = sum(norm_counts)) %>%
  distinct()

oxPTMs_hl$group <- if_else(oxPTMs_hl$LeadProt %in% short_hl_proteins$LeadProt, 'short hl', 'normal hl') #SHL = short half-life, NHL = normal half-life
ggplot(oxPTMs_hl, aes(x = mean_hl_hours, y = sum_oxPTMs, colour = group)) +
  geom_point() +
  theme_classic()

ggplotly()
```

What are the protein in the bottom part?

```{r}
oxPTMs_hl <- human_ptms_hl %>% 
  inner_join(oxPTMs, by = join_by(unimod_id == ID)) %>%
  group_by(LeadProt) %>%
  summarise(mean_hl_hours, sum_oxPTMs = sum(norm_counts)) %>%
  distinct()

unmodified <- oxPTMs_hl %>% filter(sum_oxPTMs <3) %>% dplyr::select(LeadProt)

iupred_scores_ratios <- read.csv(file = '/Users/anastasialinchik/Desktop/Thesis/DATA/generated/iupred_scores_with_ratio.tsv', header = T, sep = '\t')

df2 <- iupred_scores_ratios %>% filter(UniProt_ID %in% unmodified$LeadProt)

hist(df2$Ratio_Above_0.5)

df3 <- mt %>% filter(Protein_ID %in% unmodified$LeadProt)

hist(df3$mean_mt)
```

Colour based on stability

```{r}
oxPTMs_hl <- human_ptms_hl %>% 
  inner_join(oxPTMs, by = join_by(unimod_id == ID)) %>%
  group_by(LeadProt) %>%
  summarise(mean_hl_hours, sum_oxPTMs = sum(norm_counts)) %>%
  distinct() %>% inner_join(mt, by = join_by(LeadProt == Protein_ID))

oxPTMs_hl$group <- if_else(oxPTMs_hl$mean_mt > 55, '> 55', '< 55') #SHL = short half-life, NHL = normal half-life
ggplot(oxPTMs_hl, aes(x = mean_hl_hours, y = sum_oxPTMs, colour = group)) +
  geom_point() +
  theme_classic()

ggplotly()
```

Now look at the melting temp:

```{r}
oxPTMs_hl <- human_ptms_hl %>% 
  inner_join(oxPTMs, by = join_by(unimod_id == ID)) %>%
  group_by(LeadProt) %>%
  summarise(mean_hl_hours, sum_oxPTMs = sum(norm_counts)) %>%
  distinct() %>% inner_join(mt, by = join_by(LeadProt == Uniprot_ID))

oxPTMs_hl$group <- if_else(oxPTMs_hl$Ratio_Above_0.5 > 0.5, '>0.5', '<0.5') #SHL = short half-life, NHL = normal half-life
ggplot(oxPTMs_hl, aes(x = mean_hl_hours, y = sum_oxPTMs, colour = group)) +
  geom_point() +
  theme_classic()

ggplotly()
```

## PTMs in long-lived proteins

Long-lived proteins can be used as estimators of chronological age. Long-lived proteins can be defined in different ways, for example based on the half-life of the protein when compared to the average half-life of proteins in the organism. In this case, long-lived proteins were obtained from the following study: [paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3788602/). Proteins were classified as long-lived based on their degree of degradation during the experiment and therefore it was possible to discover new long-lived proteins (no *a priori* assumptions were made).

The study identified a list of long-lived proteins in rats, therefore human orthologs of these proteins were found.

```{r}
#|echo: false

long_lived_proteins <- read_excel(path = '/Users/anastasialinchik/Desktop/Thesis/DATA/long_lived_proteins.xlsx') %>% subset(select = c(3,4))
long_lived_proteins <- na.omit(long_lived_proteins)
long_lived_proteins_hl <- inner_join(human_ptms, long_lived_proteins, by = join_by(LeadProt == LeadProt))

human_ptms_hl_long_proteins <- rbind(long_lived_proteins_hl, human_ptms_hl)

human_ptms_hl_long_proteins$group <- if_else(human_ptms_hl_long_proteins$LeadProt %in% long_lived_proteins$LeadProt, 'long lived', 'non-long lived')
```

Just the long-lived proteins.

```{r}
ggplot(long_lived_proteins_hl, aes(x = mean_hl_hours, weight = norm_counts)) +
  geom_histogram(bins = 50) +
  labs(title = 'Distribution of half-lives of oxidised proteins', x = 'Mean half-life (hours)', y = 'Counts')+
  theme_classic()
```

```{r}
#|echo: false
human_ptms_hl_oxi <- long_lived_proteins_hl %>% filter(unimod_id %in% oxPTMs$ID)

ggplot(human_ptms_hl_oxi, aes(x = mean_hl_hours, weight = norm_counts)) +
  geom_histogram(bins = 50) +
  labs(title = 'Distribution of half-lives of oxidised proteins', x = 'Mean half-life (hours)', y = 'Counts')+
  theme_classic()
```

```{r}
#| echo: false
#| warning: false
oxPTMs_hl <- long_lived_proteins_hl %>% 
  group_by(LeadProt) %>%
  summarise(mean_hl_hours, sum_oxPTMs = sum(norm_counts)) %>%
  distinct()

ggplot(oxPTMs_hl, aes(x = mean_hl_hours, y = sum_oxPTMs)) +
  geom_point() +
  geom_smooth(method = 'lm', se = F)
  theme_classic()

ggplotly()
```

look at the 114 protein CO6A3_HUMAN

```{r}
human_ptms %>% filter(Uniprot_entry_name == 'CO6A3_HUMAN') %>%
  group_by(project_id) %>% summarise(counts = sum(norm_counts)) %>% arrange(desc(counts))
  ggplot(aes(y = counts, x = project_id)) +
  geom_bar(stats = 'identity')
```

### Long-lived proteins + the rest

The plot below shows the distribution of the half-lives of proteins with oxPTMs (normalised counts were used). Clearly there is a much greater number of short lived proteins compared to long-lived proteins.

```{r}
#|echo: false
human_ptms_hl_oxi <- human_ptms_hl_long_proteins %>% filter(unimod_id %in% oxPTMs$ID)

ggplot(human_ptms_hl_oxi, aes(x = mean_hl_hours, weight = norm_counts)) +
  geom_histogram(bins = 50) +
  labs(title = 'Distribution of half-lives of oxidised proteins', x = 'Mean half-life (hours)', y = 'Counts')+
  theme_classic()
```

### oxPTMs

This plot is similar to the one already seen above, however this time, long-lived proteins are included. There is still no correlation between the abundance of PTMs and protein half-lives.

```{r}
#| echo: false
#| warning: false
oxPTMs_hl <- human_ptms_hl_oxi %>%
  group_by(LeadProt) %>%
  summarise(group, mean_hl_hours, sum_oxPTMs = sum(norm_counts)) %>%
  distinct()


ggplot(oxPTMs_hl, aes(x = mean_hl_hours, y = sum_oxPTMs, colour = group)) +
  geom_point() +
  theme_classic()
```
