---
title: "Melting temperatures"
format: 
  html:
    embed-resources: true
    theme: flatly
    toc: true
    toc-depth: 3
    toc-expand: 3
    toc-title: Contents
    smooth-scroll: true
editor: visual
---

```{r}
#| echo: false
#| include: false

library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
library(stringr)
library(scales)
library(survey)
```

```{r}
#| echo: false
#| include: false
#human_complete <- read.csv(file = "/Users/anastasialinchik/Desktop/Thesis/DATA/generated/human_complete.tsv", header = T, sep = '\t')

#leadprot_uniprot <- human_complete %>% dplyr::select(Uniprot_entry_name, LeadProt) %>% distinct()

#THIS LOADS THE NEW DATASET
human_complete <- read.csv(file = "/Users/anastasialinchik/Desktop/Thesis/DATA/generated/human_new.tsv", header = T, sep = '\t')

human_ptms <- human_complete %>% filter(ptm_class == 'Post-translational')

oxPTMs <- read_excel(path = '/Users/anastasialinchik/Desktop/Thesis/table_oxPTMs.xlsx', sheet = 'oxPTMs')
oxPTMs <- oxPTMs %>%
  mutate(AAs_list = str_split(Specificity, ", ")) %>% dplyr::select(`Unimod #`, AAs_list) %>% rename(unimod_id = `Unimod #`)

mt <- read_xlsx(path = '/Users/anastasialinchik/Desktop/Thesis/DATA/MeltingTemperatures.xlsx', sheet = 5,  col_names = T)

mt <- separate_rows(mt, Protein_ID, sep = ";")
mt <- mt %>% dplyr::select(Protein_ID, `Tm Protein`) %>% group_by(Protein_ID) %>% summarise(mean_mt = mean(`Tm Protein`)) %>% rename('LeadProt' = 'Protein_ID')

ptms_mt <- inner_join(human_ptms, mt, by = join_by(LeadProt))
complete_mt <- inner_join(human_complete, mt, by = join_by(LeadProt))
lysine_acylation <- c(122, 1, 58, 1289, 747, 64)
AGE <- c(6, 907, 378)
```

# Initial distribution

The melting temperatures were obtained from:

Langenberg, Tobias, et al. 'Thermodynamic and Evolutionary Coupling between the Native and Amyloid State of Globular Proteins'. *Cell Reports*, vol. 31, no. 2, Apr. 2020, p. 107512. *DOI.org (Crossref)*, https://doi.org/10.1016/j.celrep.2020.03.076.

It is better to plot these as 2 overlaying density plots.

```{r}
#| echo: false
#| warning: false
ggplot() +
  geom_density(data = mt, aes(x = mean_mt, fill = 'Data found by Langenberg et al.'), alpha = 0.7, bw = 0.6) +
  geom_density(data = ptms_mt, aes(x = mean_mt, weight = counts_norm_abund_len, fill = 'Re-processed dataset'), alpha = 0.7, bw = 0.6) + 
  labs(x = 'Melting temperature (°C)', y = 'Density') +
  theme_classic() +
  scale_fill_manual(values = c("white", '#919191'),breaks = c('Data found by Langenberg et al.', 'Re-processed dataset')) +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 13, family = "Lato"),
        legend.position = c(0.28,0.9), 
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

## Outliers

```{r}
#| echo: false
ptms_mt %>% group_by(LeadProt) %>% 
  summarise(sum = log10(sum(counts_norm_abund_len))) %>%
  ggplot(aes(sum)) +
  geom_boxplot() +
  labs(x = 'Log of the sum of normalised counts')+
  theme_classic() +
  theme(legend.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.text = element_text(size = 13, family = "Lato"),
        legend.position = c(0.8,0.9), 
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

```{r}
#| echo: false
data <- ptms_mt %>% group_by(Uniprot_entry_name) %>% 
  summarise(sum_log = log10(sum(counts_norm_abund_len)))

# Calculate Q1, Q3, and IQR
Q1 <- quantile(data$sum_log, 0.25)
Q3 <- quantile(data$sum_log, 0.75)
IQR <- Q3 - Q1

# Define lower and upper bounds
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter outliers
outliers <- data %>%
  filter(sum_log > upper_bound)

ptms_mt <- ptms_mt %>% filter(!Uniprot_entry_name %in% outliers$Uniprot_entry_name)
```

Check that the outliers have been removed.

```{r}
#| echo: false
ptms_mt %>% group_by(LeadProt) %>% 
  summarise(sum = log10(sum(counts_norm_abund_len))) %>%
  ggplot(aes(sum)) +
  geom_boxplot() +
  labs(x = 'Log of the sum of normalised counts')+
  theme_classic() +
  theme(legend.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.text = element_text(size = 13, family = "Lato"),
        legend.position = c(0.8,0.9), 
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

## Distributions

### Phoshorylation

```{r}
#| echo: false
#| warning: false
human_ptms_hl_mod_group <- ptms_mt
human_ptms_hl_mod_group$mod_group <- if_else(human_ptms_hl_mod_group$unimod_id == 21, 'Phosphorylated', 'Non-phosphorylated')

human_ptms_hl_mod_group$mod_group <- factor(human_ptms_hl_mod_group$mod_group, levels = c("Non-phosphorylated", "Phosphorylated"))

# all of the proteins that get phosphorylated
pho_prot <- human_ptms_hl_mod_group %>% filter(mod_group == 'Phosphorylated') %>% dplyr::select(LeadProt) %>% distinct()
# those proteins that get phosphorylated are not included in the Non-phosphorylated section

human_ptms_hl_mod_group_exclusive <- human_ptms_hl_mod_group %>% filter(mod_group == 'Phosphorylated' | (mod_group == 'Non-phosphorylated' & (!LeadProt %in% pho_prot$LeadProt)))

p1 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Non-phosphorylated') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 
p2 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Phosphorylated') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 

proteins_exclusive <- rbind(p1, p2)

ggplot() +
  geom_violin(data = human_ptms_hl_mod_group_exclusive, aes(x = mod_group, y = mean_mt, weight = counts_norm_abund_len, fill = mod_group)) +
  geom_boxplot(data = proteins_exclusive, aes(x = mod_group, y = mean_mt, weight = sum), width = 0.1,  alpha = 0.5, position = position_dodge(width = 0.5)) +
  scale_fill_manual(values = c("white", '#b3de69')) +
  labs(y = 'Mean half-lives')+
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = 'none',
        axis.title.x = element_blank(),
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

Wilcoxon test (note that the sample sizes are uneven). The p value was adjusted uisng the formula p/sqrt(N/100), where N = n1+ n2

```{r}
#| echo: false
# Create survey design
design <- svydesign(ids = ~1, data = proteins_exclusive, weights = ~sum)

# Perform the weighted Wilcoxon test using svyranktest
svyranktest(mean_mt~mod_group, design, test = "wilcoxon")

print('P-value cut-off')
0.05/sqrt((dim(proteins_exclusive)[1])/100)
```

### Acetylation

```{r}
#| echo: false
#| warning: false
#| message: false
human_ptms_hl_mod_group <- ptms_mt 
human_ptms_hl_mod_group$mod_group <- if_else(human_ptms_hl_mod_group$unimod_id == 1, 'Acetylated', 'Non-acetylated')

human_ptms_hl_mod_group$mod_group <- factor(human_ptms_hl_mod_group$mod_group, levels = c("Non-acetylated", "Acetylated"))

# all of the proteins that get acetylated
acetyl_prot <- human_ptms_hl_mod_group %>% filter(mod_group == 'Acetylated') %>% dplyr::select(LeadProt) %>% distinct()
# those proteins that get methylated are not included in the non-meth section
human_ptms_hl_mod_group_exclusive <- human_ptms_hl_mod_group %>% filter(mod_group == 'Acetylated' | (mod_group == 'Non-acetylated' & (!LeadProt %in% acetyl_prot$LeadProt)))

p1 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Non-acetylated') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 
p2 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Acetylated') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 

proteins_exclusive <- rbind(p1, p2)

ggplot() +
  geom_violin(data = human_ptms_hl_mod_group_exclusive, aes(x = mod_group, y = mean_mt, weight = counts_norm_abund_len, fill = mod_group)) +
  geom_boxplot(data = proteins_exclusive, aes(x = mod_group, y = mean_mt, weight = sum), width = 0.1,  alpha = 0.5, position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("white", '#8dd3c7')) +
  labs(y = 'Mean half-lives')+
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = 'none',
        axis.title.x = element_blank(),
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

```{r}
#| echo: false
# Create survey design
design <- svydesign(ids = ~1, data = proteins_exclusive, weights = ~sum)

# Perform the weighted Wilcoxon test using svyranktest
svyranktest(mean_mt~mod_group, design, test = "wilcoxon")

print('P-value cut-off')
0.05/sqrt((dim(proteins_exclusive)[1])/100)
```

### Ubiquitination

```{r}
#| echo: false
#| warning: false
#| message: false
ubi_prot <- human_complete %>% filter(unimod_id == 121) %>% inner_join(mt, by = join_by(LeadProt))

human_ptms_hl_mod_group <- ptms_mt
human_ptms_hl_mod_group <- rbind(human_ptms_hl_mod_group, ubi_prot)

human_ptms_hl_mod_group$mod_group <- if_else(human_ptms_hl_mod_group$unimod_id == 121, 'Ubiquitinated', 'Non-ubiquitinated')

human_ptms_hl_mod_group$mod_group <- factor(human_ptms_hl_mod_group$mod_group, levels = c("Non-ubiquitinated", "Ubiquitinated"))

# all of the proteins that get oxidised
ubi_prot <- human_ptms_hl_mod_group %>% filter(mod_group == 'Ubiquitinated') %>% dplyr::select(LeadProt) %>% distinct()
# those proteins that get oxidised are not included in the non-oxi section
human_ptms_hl_mod_group_exclusive <- human_ptms_hl_mod_group %>% filter(mod_group == 'Ubiquitinated' | (mod_group == 'Non-ubiquitinated' & (!LeadProt %in% ubi_prot$LeadProt)))

p1 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Non-ubiquitinated') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 
p2 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Ubiquitinated') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 

proteins_exclusive <- rbind(p1, p2)

ggplot() +
  geom_violin(data = human_ptms_hl_mod_group_exclusive, aes(x = mod_group, y = mean_mt, weight = counts_norm_abund_len, fill = mod_group)) +
  geom_boxplot(data = proteins_exclusive, aes(x = mod_group, y = mean_mt, weight = sum), width = 0.1,  alpha = 0.5, position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("white", '#bebada')) +
  labs(y = 'Mean half-lives')+
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = 'none',
        axis.title.x = element_blank(),
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

```{r}
#| echo: false
# Create survey design
design <- svydesign(ids = ~1, data = proteins_exclusive, weights = ~sum)

# Perform the weighted Wilcoxon test using svyranktest
svyranktest(mean_mt~mod_group, design, test = "wilcoxon")

print('P-value cut-off')
0.05/sqrt((dim(proteins_exclusive)[1])/100)
```

### Methylation

```{r}
#| echo: false
#| warning: false
#| message: false
human_ptms_hl_mod_group <- ptms_mt
human_ptms_hl_mod_group$mod_group <- if_else(human_ptms_hl_mod_group$unimod_id == 34, 'Methylated', 'Non-methylated')
human_ptms_hl_mod_group <- human_ptms_hl_mod_group 

human_ptms_hl_mod_group$mod_group <- factor(human_ptms_hl_mod_group$mod_group, levels = c("Non-methylated", "Methylated"))

# all of the proteins that get methylated
meth_prot <- human_ptms_hl_mod_group %>% filter(mod_group == 'Methylated') %>% dplyr::select(LeadProt) %>% distinct()
# those proteins that get methylated are not included in the non-meth section
human_ptms_hl_mod_group_exclusive <- human_ptms_hl_mod_group %>% filter(mod_group == 'Methylated' | (mod_group == 'Non-methylated' & (!LeadProt %in% meth_prot$LeadProt)))

p1 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Non-methylated') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 
p2 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Methylated') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 

proteins_exclusive <- rbind(p1, p2)

ggplot() +
  geom_violin(data = human_ptms_hl_mod_group_exclusive, aes(x = mod_group, y = mean_mt, weight = counts_norm_abund_len, fill = mod_group)) +
  geom_boxplot(data = proteins_exclusive, aes(x = mod_group, y = mean_mt, weight = sum), width = 0.1,  alpha = 0.5, position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("white", '#80b1d3')) +
  labs(y = 'Mean half-lives')+
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = 'none',
        axis.title.x = element_blank(),
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

```{r}
#| echo: false
# Create survey design
design <- svydesign(ids = ~1, data = proteins_exclusive, weights = ~sum)

# Perform the weighted Wilcoxon test using svyranktest
svyranktest(mean_mt~mod_group, design, test = "wilcoxon")

print('P-value cut-off')
0.05/sqrt((dim(proteins_exclusive)[1])/100)
```

### oxPTMs

```{r}
#| echo: false
#| warning: false
#| message: false

human_ptms_hl_mod_group <- complete_mt 

human_ptms_hl_mod_group$mod_group <- if_else((human_ptms_hl_mod_group$unimod_id %in% oxPTMs$unimod_id & human_ptms_hl_mod_group$ptm_res %in% oxPTMs$AAs_list), 'oxPTMs', 'Non-oxPTMs')

human_ptms_hl_mod_group$mod_group <- factor(human_ptms_hl_mod_group$mod_group, levels = c("Non-oxPTMs", "oxPTMs"))

human_ptms_hl_mod_group_exclusive <- human_ptms_hl_mod_group %>% filter((mod_group == 'Non-oxPTMs' & ptm_class == 'Post-translational') | mod_group == 'oxPTMs')

p1 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Non-oxPTMs') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 
p2 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'oxPTMs') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 

proteins_exclusive <- rbind(p1, p2)

ggplot() +
  geom_violin(data = human_ptms_hl_mod_group_exclusive, aes(x = mod_group, y = mean_mt, weight = counts_norm_abund_len, fill = mod_group)) +
  geom_boxplot(data = proteins_exclusive, aes(x = mod_group, y = mean_mt, weight = sum), width = 0.1,  alpha = 0.5, position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("white", '#fdb462')) +
  labs(y = 'Mean half-lives')+
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = 'none',
        axis.title.x = element_blank(),
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

```{r}
#| echo: false
# Create survey design
design <- svydesign(ids = ~1, data = proteins_exclusive, weights = ~sum)

# Perform the weighted Wilcoxon test using svyranktest
svyranktest(mean_mt~mod_group, design, test = "wilcoxon")

print('P-value cut-off')
0.05/sqrt((dim(proteins_exclusive)[1])/100)
```

### Lysine acylations

```{r}
#| echo: false
#| warning: false
#| message: false
human_ptms_hl_mod_group <- complete_mt 

human_ptms_hl_mod_group$mod_group <- if_else((human_ptms_hl_mod_group$unimod_id %in% lysine_acylation & human_ptms_hl_mod_group$ptm_res == 'K'), 'K acylation', 'Non-acylated')

human_ptms_hl_mod_group$mod_group <- factor(human_ptms_hl_mod_group$mod_group, levels = c("Non-acylated", "K acylation"))

human_ptms_hl_mod_group_exclusive <- human_ptms_hl_mod_group %>% filter((mod_group == 'Non-acylated' & ptm_class == 'Post-translational') | mod_group == 'K acylation')

p1 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Non-acylated') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 
p2 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'K acylation') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 

proteins_exclusive <- rbind(p1, p2)

ggplot() +
  geom_violin(data = human_ptms_hl_mod_group_exclusive, aes(x = mod_group, y = mean_mt, weight = counts_norm_abund_len, fill = mod_group)) +
  geom_boxplot(data = proteins_exclusive, aes(x = mod_group, y = mean_mt, weight = sum), width = 0.1,  alpha = 0.5, position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("white", '#fb8072')) +
  labs(y = 'Mean half-lives')+
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = 'none',
        axis.title.x = element_blank(),
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

```{r}
#| echo: false
# Create survey design
design <- svydesign(ids = ~1, data = proteins_exclusive, weights = ~sum)

# Perform the weighted Wilcoxon test using svyranktest
svyranktest(mean_mt~mod_group, design, test = "wilcoxon")

print('P-value cut-off')
0.05/sqrt((dim(proteins_exclusive)[1])/100)
```

### AGEs

```{r}
#| echo: false
#| warning: false
#| message: false
human_ptms_hl_mod_group <- complete_mt 

human_ptms_hl_mod_group$mod_group <- if_else((human_ptms_hl_mod_group$unimod_id %in% AGE & human_ptms_hl_mod_group$ptm_res == 'K'), 'AGEs', 'Non-AGEs')

human_ptms_hl_mod_group$mod_group <- factor(human_ptms_hl_mod_group$mod_group, levels = c("Non-AGEs", "AGEs"))

human_ptms_hl_mod_group_exclusive <- human_ptms_hl_mod_group %>% filter((mod_group == 'Non-AGEs' & ptm_class == 'Post-translational') | mod_group == 'AGEs')

p1 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'Non-AGEs') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 
p2 <- human_ptms_hl_mod_group_exclusive %>% filter(mod_group == 'AGEs') %>% group_by(Uniprot_entry_name) %>% summarise(sum =sum(counts_norm_abund_len), mean_mt, mod_group) %>%
distinct(Uniprot_entry_name, mean_mt, mod_group, sum) 

proteins_exclusive <- rbind(p1, p2)

ggplot() +
  geom_violin(data = human_ptms_hl_mod_group_exclusive, aes(x = mod_group, y = mean_mt, weight = counts_norm_abund_len, fill = mod_group)) +
  geom_boxplot(data = proteins_exclusive, aes(x = mod_group, y = mean_mt, weight = sum), width = 0.1,  alpha = 0.5, position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("white", '#fccde5')) +
  labs(y = 'Mean half-lives')+
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = 'none',
        axis.title.x = element_blank(),
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

```{r}
#| echo: false
# Create survey design
design <- svydesign(ids = ~1, data = proteins_exclusive, weights = ~sum)

# Perform the weighted Wilcoxon test using svyranktest
svyranktest(mean_mt~mod_group, design, test = "wilcoxon")

print('P-value cut-off')
0.05/sqrt((dim(proteins_exclusive)[1])/100)
```

# Binning by MT

```{r}
#| echo: false
#| warning: false
#| message: false

df <- ptms_mt %>%
  mutate(mt_group = case_when(
    mean_mt <= 45 ~ "40-45",
    mean_mt <= 50 ~ "45-50",
    mean_mt <= 55 ~ "50-55",
    mean_mt <= 60 ~ "55-60",
    mean_mt <= 65 ~ "60-65",
    mean_mt <= 70 ~ "65-70",
    TRUE ~ "70+"
  )) %>%
    mutate(mod_group = case_when(
    unimod_id == 1 ~'Acetylated',
    unimod_id == 34 ~'Methylated',
    TRUE ~ "-"
  )) 

mean_counts_per_mt_group <- df %>%
  group_by(Uniprot_entry_name, mod_group) %>%
  summarize(sum_ptms = sum(counts_norm_abund_len), mt_group) %>% 
  filter(!mod_group == '-') %>%
  distinct() %>% group_by(mod_group, mt_group) %>% summarise(mean_ptms_group = mean(sum_ptms))

mod_group_colours <- c('Acetylated' = '#8dd3c7', 'Methylated' = '#80b1d3')

# Plot bar chart
ggplot(mean_counts_per_mt_group, aes(x = mt_group, y = mean_ptms_group, fill = mod_group)) +
  geom_bar(stat = "identity", position = 'dodge') +
  scale_x_discrete(limits = c("40-45", "45-50", "50-55", "55-60", "60-65", "65-70")) +
  scale_y_continuous(limits = c(0, 0.0004), labels = scales::label_number(scale = 1e4)) +
  scale_fill_manual(values = mod_group_colours) +
  labs(x = "Melting temperature (°C)",
       y = "Mean normalised counts (scaled by 1e4)") +
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 13, family = "Lato"),
        legend.position = c(0.8,0.92), 
        legend.direction = 'horizontal',
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

```{r}
#| echo: false
#| warning: false
#| message: false
ubi_prot <- human_complete %>% filter(unimod_id == 121) %>% inner_join(mt, by = join_by(LeadProt))
df <- rbind(ptms_mt, ubi_prot)

a <- human_complete %>% filter(unimod_id %in% c(lysine_acylation, AGE)) %>% inner_join(mt, by = join_by(LeadProt))
df <- rbind(df, a)

df <- df %>%
  mutate(mt_group = case_when(
    mean_mt <= 45 ~ "40-45",
    mean_mt <= 50 ~ "45-50",
    mean_mt <= 55 ~ "50-55",
    mean_mt <= 60 ~ "55-60",
    mean_mt <= 65 ~ "60-65",
    mean_mt <= 70 ~ "65-70",
    TRUE ~ "70+"
  )) %>%
  mutate(mod_group = case_when(
    unimod_id == 121 ~'Ubiquitinated',
    unimod_id %in% AGE ~ 'AGE',
    (unimod_id %in% lysine_acylation & ptm_res == 'K') ~ 'K acylation',
    TRUE ~ "-"
  ))

mean_counts_per_mt_group <- df %>%
  group_by(Uniprot_entry_name, mod_group) %>%
  summarize(sum_ptms = sum(counts_norm_abund_len), mt_group) %>% 
  filter(!mod_group == '-') %>%
  distinct() %>% group_by(mod_group, mt_group) %>% summarise(mean_ptms_group = mean(sum_ptms))

mod_group_colours <- c('AGE' = '#fccde5', 'Ubiquitinated' = '#bebada', 'K acylation' = '#fb8072')

# Plot bar chart
ggplot(mean_counts_per_mt_group, aes(x = mt_group, y = mean_ptms_group, fill = mod_group)) +
  geom_bar(stat = "identity", position = 'dodge') +
  scale_x_discrete(limits = c("40-45", "45-50", "50-55", "55-60", "60-65", "65-70")) +
  scale_y_continuous(limits = c(0,0.00035), labels = scales::label_number(scale = 1e4)) +
  scale_fill_manual(values = mod_group_colours) +
  labs(x = "Melting temperature (°C)",
       y = "Mean normalised counts (scaled by 1e4)") +
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 13, family = "Lato"),
        legend.position = c(0.72,0.9), 
        legend.direction = 'horizontal',
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

```{r}
#| echo: false
#| warning: false
#| message: false

oxptms_prot <- human_complete %>% filter(unimod_id %in% oxPTMs$unimod_id) %>% inner_join(mt, by = join_by(LeadProt))
df <- rbind(ptms_mt, oxptms_prot)

df <- ptms_mt %>%
  mutate(mt_group = case_when(
    mean_mt <= 45 ~ "40-45",
    mean_mt <= 50 ~ "45-50",
    mean_mt <= 55 ~ "50-55",
    mean_mt <= 60 ~ "55-60",
    mean_mt <= 65 ~ "60-65",
    mean_mt <= 70 ~ "65-70",
    TRUE ~ "70+"
  )) %>%
    mutate(mod_group = case_when(
    unimod_id == 21 ~'Phosphorylated',
    (unimod_id %in% oxPTMs$unimod_id & ptm_res %in% oxPTMs$AAs_list) ~ 'oxPTMs',
    TRUE ~ "-"
  )) 

mean_counts_per_mt_group <- df %>%
  group_by(Uniprot_entry_name, mod_group) %>%
  summarize(sum_ptms = sum(counts_norm_abund_len), mt_group) %>% 
  filter(!mod_group == '-') %>%
  distinct() %>% group_by(mod_group, mt_group) %>% summarise(mean_ptms_group = mean(sum_ptms))

mod_group_colours <- c('Phosphorylated' = '#b3de69', 'oxPTMs' = '#fdb462')

# Plot bar chart
ggplot(mean_counts_per_mt_group, aes(x = mt_group, y = mean_ptms_group, fill = mod_group)) +
  geom_bar(stat = "identity", position = 'dodge') +
  scale_x_discrete(limits = c("40-45", "45-50", "50-55", "55-60", "60-65", "65-70")) +
  scale_y_continuous(labels = scales::label_number(scale = 1e4)) +
  scale_fill_manual(values = mod_group_colours) +
  labs(x = "Melting temperature (°C)",
       y = "Mean normalised counts (scaled by 1e4)") +
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 13, family = "Lato"),
        legend.position = c(0.75,0.93), 
        legend.direction = 'horizontal',
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"))
```

# Binning by the number of PTMs

```{r}
#| echo: false
#| message: false
#| warning: false
bins <- ptms_mt %>% group_by(Uniprot_entry_name) %>% summarise(mean_mt, sum_norm_counts = sum(counts_norm_abund_len)) %>% distinct() %>% arrange(desc(sum_norm_counts))
#dim(bins)
bin1 <- bins[1:284, ]
bin2 <- bins[285:568, ]
bin3 <- bins[569:852, ]

bin1$bin <- 'High'
bin2$bin <- 'Medium'
bin3$bin <- 'Low'

bins <- rbind(bin1, bin2, bin3)
bins$bin <- factor(bins$bin, levels = c("Low", "Medium", "High"))
ggplot(bins, aes(x = mean_mt)) +
  geom_histogram(fill = '#919191') +
  labs(x = 'Melting temperature (°C)', y = 'Protein count') +
  facet_wrap(~bin) +
  theme_bw() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        strip.text = element_text(size = 13, family = "Lato"),
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"),
        strip.background.x = element_blank()
        )
```

```{r}
#| echo: false
#| message: false

bin1 <- bins[1:94, ]
bin2 <- bins[95:189, ]
bin3 <- bins[190:284, ]

bin1$bin <- 'High'
bin2$bin <- 'Medium'
bin3$bin <- 'Low'

bins <- rbind(bin1, bin2, bin3)
bins$bin <- factor(bins$bin, levels = c("Low", "Medium", "High"))
ggplot(bins, aes(x = mean_mt)) +
  geom_histogram( fill = '#919191') +
  labs(x = 'Melting temperature (°C)', y = 'Protein count') +
  facet_wrap(~bin) +
  theme_bw() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        strip.text = element_text(size = 13, family = "Lato"),
        axis.text = element_text(size = 13, family = "Lato"),  
        axis.title = element_text(size = 13, family = "Lato"),
        strip.background.x = element_blank()
        )
```
